\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amsfonts}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}

\newcommand{\tuple}[1]{\langle #1 \rangle}
\newcommand{\concat}{\mathbin{\widehat{}}}
\newcommand{\set}[1]{\{ #1 \}}
\newcommand{\compr}[2]{\{ #1 \,:\, #2 \}}
\newcommand{\terminates}{\!\!\downarrow}
\newcommand{\diverges}{\!\!\uparrow}
\newcommand{\njump}{\emptyset^{(n)}}
\newcommand{\njumpprime}{\emptyset^{(n)^\prime}}
\newcommand{\snjump}{\emptyset^{(n+1)}}
\newcommand{\PiOne}{\Pi^0_1}
\newcommand{\Cons}{\widehat{\,\,}}
\DeclareMathOperator{\Fin}{Fin}
\DeclareMathOperator{\Tot}{Tot}
\DeclareMathOperator{\Comp}{Comp}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\use}{use}

\title{Computability Theory: Week 14}
\author{Anton Golov (agolov@science.ru.nl)}
\begin{document}
  \maketitle

  Exercise: 6.3.2.

  \begin{theorem}
    There exists a low simple set $A$.
  \end{theorem}

  \begin{proof}
    We build the set $A$ using a finite injury construction.  We have the following requirements:
    \begin{align*}
      R_e &= \text{$W_e$ infinite} \implies A \cap W_e \neq \emptyset\\
      L_e &= \exists^\infty s.\, \set{e}^{A_s}_s(e)\terminates \implies \set{e}^A(e)\terminates.
    \end{align*}

    The $R_e$ requirements ensure that $A$ is simple, while the $L_e$ requirements ensure $A$ is low.  The priority
    ordering on these requirements is
    \[
      R_0 > L_0 > R_1 > L_1 > \cdots
    \]

    Throughout the construction, we maintain a restraint function $r(e, s)$ that will ensure that we do not injure
    higher priority requirements.  If at stage $s+1$ we do not specify a value for $r(e, s+1)$ we mean that the value
    from the previous stage should be maintained, that is, that that we set $r(e, s+1) = r(e,s)$.

    We say that a requirement $R_e$ requires attention at stage $s+1$ if $W_{e,s} \cap A_s = \emptyset$ and there is an
    $x \in W_{e,s}$ such that $\forall d < e$, $r(d, s) < x$ and $2e < x$.  We say a requirement $L_e$ requires
    attention at stage $s+1$ if $\set{e}^{A_s}_s(e)\terminates$ and $r(e, s) < \use(\set{e}^{A_s}_s(e))$.

    The construction proceeds as follows.

    Stage $s = 0$: Set $A_0 = \emptyset$ and set $r(e, 0) = 0$ for all $e$.

    Stage $s+1$: Let $d < s$ be minimal such that $R_d$ or $L_d$ requires attention.  If no such $d$ exists, do nothing
    and proceed to the next stage.

    If $R_d$ requires attention, then there is some $x \in W_{e,s}$ such that for all $d < e$, $r(d, s) < x$ and $2e <
    x$.  Enumerate $x$ into $A$.

    If, instead, $L_d$ requires attention, then set $r(d, s+1) = \use(\set{d}^{A_s}_s(d))$.  This computation converges
    since $L_d$ requires attention.  This concludes the construction.

    We verify that the requirements are satisfied.  Let $e$ be such that $W_e$ is infinite.  By induction, there is some
    stage $s$ such that all requirements with priority higher than $R_e$ will not act again.  Since $W_e$ is infinite,
    there is some least stage $s'$ and $x \in W_{e,s'}$ such that for all $d < e$, $r(d, s) < x$, and $2e < x$.  It
    follows that $R_e$ will eventually require attention and thus $A \cap W_e \neq \emptyset$ since $x$ will be
    enumerated into $A$.

    Now let $e$ be such that $\exists^\infty s.\, \set{e}^{A_s}_s(e)\terminates$.  Again, by induction let $s$ be a
    stage such that all higher priority requirements than $L_e$ will not act after $s$.  At some stage $s' > s$,
    $\set{e}^{A_{s'}}_{s'}(e)\terminates$ with use $u$ and thus $L_e$ acts and $r(c, s'+1)$ is set to $u$.  Since the
    restraint functions are non-decreasing in the second parameter, no value below $u$ will be enumerated into $A$ from
    stage $s'$ onwards and so $\set{e}^{A}(e)\terminates$.

    Note that every $R_e$ requirements acts at most once, since once an element of $W_{e,s}$ is enumerated into $A$, for
    all $s' > s$ we have $W_{e,s'} \cap A_{s'} \neq \emptyset$.  Every $L_e$ requirement acts at most once after it is
    injured by a higher priority requirement; by induction, every $L_e$ acts at most finitely often.

    By Exercise~5.6.10 it follows that $A$ is low.  Note that $\bar{A}$ is infinite: every requirement $R_e$ enumerates
    at most one value into $A$, and this value is above $2e$, while the requirements $L_e$ do not enumerate anything
    into $A$.  It follows that $|\compr{x \in A}{x \le 2e}| \le e$.  Finally, note that for every infinite c.e.\@ set
    $W_e$, since $W_e \cap A \neq \emptyset$, $W_e$ is not a subset of $\bar{A}$ and thus $A$ is simple.
  \end{proof}
\end{document}
